\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{titlesec}

\setstretch{1.1}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\begin{document}

\begin{center}
    {\Large \textbf{Graph Neural Networks for CT-Based Representation Learning and Similarity Mapping of Known and Unknown Lung Nodule Patterns}}\\[1.5em]
    
    Swathi Gopal\\
    Harrison Lavins\\
    Rensildi Kalanxhi\\[1em]
    
    CSC 7760 Deep Learning\\
    Professor Ming Dong\\
    April 6, 2026
\end{center}

\section*{Abstract}

This project investigates whether Graph Neural Networks (GNNs) can learn meaningful relationships among lung nodules in CT scans and use those learned relationships to analyze how uncertain or previously unseen patterns relate to known nodule groups. Using the publicly available LIDC-IDRI dataset, each lung nodule is represented as a node in a graph, while edges connect nodules that share similar imaging features. The main objective is to learn graph-based representations of known lung nodule patterns and examine whether the learned embedding space can support similarity-based analysis of uncertain cases. By focusing only on a GNN-based framework, this project provides a simple and realistic approach for studying graph-based representation learning in lung CT analysis.

\section*{Goal}

The goal of this project is to use a Graph Neural Network to learn feature relationships among lung nodules from CT scans and classify known patterns while examining how unknown or uncertain patterns relate to the learned groups.

\section*{Background}

Deep learning methods have shown strong performance in medical imaging, but many models treat each sample independently and do not explicitly model relationships between similar cases. Graph Neural Networks provide a way to represent each lung nodule as part of a larger graph structure, where nodules with similar imaging features can influence one another. This makes GNNs a useful approach for learning patterns in lung CT data. In this project, the graph-based framework is used to capture relationships among nodules and to study whether uncertain or weakly matched nodules can be analyzed relative to the known groups through the learned embedding space.

\section*{Deep Learning Model}

The main model used in this project is a \textbf{Graph Neural Network}, specifically \textbf{GraphSAGE}. Each node in the graph represents a lung nodule, and edges connect nodules that are similar based on extracted CT imaging features. The GNN learns embeddings for each node and uses these embeddings for classification and similarity analysis. The focus of the project is intentionally limited to the GNN model in order to keep the scope realistic and clear.

\section*{Experiments Planned}

The experiment begins by constructing a graph from lung nodules in the LIDC-IDRI dataset. A GNN will then be trained to classify known nodule patterns based on the graph structure and learned node features. After training, the learned embeddings will be examined to determine how uncertain or weakly matched nodules relate to the known groups in the embedding space. This allows the project to explore both classification and similarity-based representation learning in a simple and focused setting.

\section*{Dataset}

The dataset used in this project is \textbf{LIDC-IDRI}, a public lung CT dataset containing thoracic CT scans and annotated lung nodules. This dataset is suitable for the project because it provides lesion-level information that can be used to construct a graph and train a Graph Neural Network model. Since CT data captures lesion and nodule structure rather than individual cancer cells, the unit of analysis in this study is the lung nodule.

\section*{References}

\begin{enumerate}[label={[\arabic*]}]
    \item Zhang, J., Ding, Y., \& Mazurowski, M. A. (2020). \textit{Towards safe and reliable deep learning for lung nodule malignancy estimation using out-of-distribution detection}.
    \item Cordella, L. P., Foggia, P., Sansone, C., \& Vento, M. (2004). \textit{A SubGraph Isomorphism Algorithm for Matching Large Graphs}. IEEE Transactions on Pattern Analysis and Machine Intelligence.
    \item LIDC-IDRI Dataset.
\end{enumerate}

\end{document}